---
title: "Compute phenotypes"
author: "Jarkko Toivonen"
date: "2/16/2021"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages(library(tidyverse))
library(lubridate)
library(readxl)
library(RVenn)
library("UpSetR")
library(knitr)
library(magrittr)  # for set_colnames

source("common.R")

# This can measure time a chunk took to execute.
# Add chunk option time_it=TRUE to each chunk your want to measure.
knitr::knit_hooks$set(time_it = local({
  now <- NULL
  function(before, options) {
    if (before) {
      # record the current time before each chunk
      now <<- lubridate::now()
    } else {
      # calculate the time difference after a chunk
      res <- lubridate::now() - now
      # return a character string to show the time
      msg <- paste("Time for this code chunk to run:", as.numeric(res), units(res))
      message(msg)
      NULL   # Don't return the message so that it won't be printed to the resulting document.
    }
  }
}))


knitr::opts_chunk$set(time_it=TRUE)

threshold <- as_date("2019-12-31")   # Don't take donations newer than this
write_results <- TRUE

result_dir <- '~/FRCBS/blood_health_phewas/results'
```




```{r Functions}




get_deferral <- function() {
  deferral_filename <- "~/proj/interval_prediction/data/FRC.DW_DEFERRAL.DAT"
  input_col_types <- list(
    X1 = col_character(),
    X2 = col_factor(),
    X3 = col_date("%Y%m%d"),
    X4 = col_date("%Y%m%d")
  )
  deferral <- read_delim(deferral_filename, delim="|", col_names = F, col_types = input_col_types)
  
  names(deferral) <- c("KEY_DONOR",
                       "KEY_DEFERRAL_CODE",
                       "KEY_DEFERRAL_DATE_START",
                       "KEY_DEFERRAL_DATE_STOP")
  
  deferral <- 
    deferral %>%
    rename(donor = KEY_DONOR)
  # Filter out donors with unrealistic deferral dates
  old_count <- nrow(deferral); old_count2 <- ndonor(deferral)
  deferral <- deferral %>%
    filter(KEY_DEFERRAL_DATE_START > "1909-10-22") %>%
    filter(KEY_DEFERRAL_DATE_START <= KEY_DEFERRAL_DATE_STOP)
  cat(sprintf("dw_deferral: Dropped %i / %i deferrals (%i / %i donors) because of date older than 1909-10-22 or start_date > stop date\n",
              old_count - nrow(deferral), old_count, old_count2 - ndonor(deferral), old_count2))
  
  # If there are multiple stop dates for a deferral, choose the latest
  old_count <- nrow(deferral); old_count2 <- ndonor(deferral)
  deferral <- deferral %>%
    group_by(donor, KEY_DEFERRAL_CODE, KEY_DEFERRAL_DATE_START) %>%
    slice_max(order_by=KEY_DEFERRAL_DATE_STOP, n=1) %>%
    ungroup()
  cat(sprintf("dw_deferral: Dropped %i / %i deferrals (%i / %i donors) because of multiple deferral stop dates\n",
              old_count - nrow(deferral), old_count, old_count2 - ndonor(deferral), old_count2))
  
  return(deferral)
}

get_donation <- function(donation.file) {
  # In the full dataset there lots of missing values. This causes automatic recognition of column types to fail.
  # Therefore we give them explicitly here.
  input_col_types <- list(
    X1 = col_character(),
    X2 = col_character(),
    X3 = col_character(),
    X4 = col_character(),
    X5 = col_character(),
    X6 = col_character(),
    X7 = col_character(),
    X8 = col_character(),
    X9 = col_character(),
    X10 = col_character(),
    X11 = col_double(),
    X12 = col_double()
  )
  
  donation <- read_delim(donation.file, col_names=FALSE, delim='|', col_types=input_col_types)
  cat(sprintf("Read %i rows from file %s\n", nrow(donation), donation.file))
  names(donation)=c('donation', 'donor', 'site', 'date', 'phleb_start',
                    'status', 'donat_phleb', 
                    'directed', 'donStartTime', 'volume_drawn', 'index_test', 
                    'Hb')
  
  old_count <- nrow(donation); old_count2 <- ndonor(donation)
  donation <- 
    donation %>%
    mutate_at(c("status", "donat_phleb", "directed", "index_test"), as.factor) %>%
    mutate_at("volume_drawn", as.integer) %>%
    mutate(date = ymd(date)) %>%
    filter(date <= threshold)
  cat(sprintf("eProgesa: Dropped %i / %i donations (%i / %i donors) because of date greater than %s\n",
              old_count - nrow(donation), old_count, old_count2 - ndonor(donation), old_count2, threshold))
   
  # Select the last attempt of the day
  old_count <- nrow(donation); old_count2 <- ndonor(donation)
  donation <- donation %>%
    group_by(donor, date) %>%
    slice_max(order_by = phleb_start, n = 1) %>%
    ungroup()
  cat(sprintf("eProgesa: Dropped %i / %i donations (%i / %i donors) because of choosing the last attempt of the day %s\n",
              old_count - nrow(donation), old_count, old_count2 - ndonor(donation), old_count2, threshold))
  
  donation <- donation %>%
    group_by(donor) %>%
    mutate(previous_Hb = lag(Hb, order_by = date)) %>%
    ungroup()
  return(donation)
}

get_donor <- function(donor_file) {
  input_col_types2 <- list(
    X1 = col_character())
  donors <- read_delim(donor_file, col_names=FALSE, delim="|", col_types = input_col_types2)
  cat(sprintf("Read %i rows from file %s\n", nrow(donors), donor_file))
  old_names <-
    c("KEY_DONOR", "DONOR_FIRST", "DONOR_NAME", "KEY_DONOR_SEX", "KEY_DONOR_DOB", "DONOR_LANGUAGE", "KEY_DONOR_ABORH", "DONOR_ADDR_1_A", 
    "KEY_DONOR_ZIP_1", "DONOR_CITY_1", "DONOR_TEL_1", "DONOR_E_MAIL", "DONOR_TEL_MOBILE", "DONOR_NOTIFIABLE", "DONOR_NOTIFICATION_METHOD_1", 
    "DONOR_NOTIFICATION_METHOD_2", "DONOR_NOTIFICATION_METHOD_3", "DONOR_NB_DONATIONS", "DONOR_NB_DONAT_PROGESA", "DONOR_NB_DONAT_OUTSIDE", 
    "DONOR_DATE_FIRST_DONATION", "DONOR_NB_WB", "DONOR_NB_PLA", "DONOR_NB_THR", "DONOR_LAST_DONAT_PHLEB", "DONOR_LAST_COLLECT")
  names(donors) <- old_names
  donors
}

get_age <- function(donor.file) {
  input_col_types2 <- list(
    X1 = col_character(),
    X5 = col_character()
    )
  donor <- read_delim(donor.file, col_names=FALSE, delim="|", col_types = input_col_types2)
  names(donor)=c('donor','first','family', 'gender', 'dob', 'language', 'aborh', 'address', 'zip', 'city',
                 'tel','email', 'mobile',
                 'notifiable', 'notification_method_1', 'notification_method_2', 'notification_method_3', 
                 'nb_donations', 'nb_donat_progesa', 'nb_donat_outside', 
                 'date_first_donation', 'nb_wb', 'nb_pla',
                 'nb_thr', 'last_donat_phleb', 'last_collect'
                 
  )
  donor <- donor %>%
    mutate(dob = ymd(dob)) %>%
    mutate(age = as.period(interval(start = dob, end = threshold))$year)
  
  return(donor %>% select(donor, age))
}



get_defects <- function() {
  vanhat <- read_excel("~/proj/interval_prediction/data/Haitat 2010-2018.xlsx", sheet="listaus")
  uudet <- read_excel("~/proj/interval_prediction/data/Haitat 2019-01-01--2020-04-30.xlsx")
  defects <- bind_rows(vanhat, uudet)
  defects <- defects %>% 
    rename(donor="Luovuttaja koodi") 
  old_count <- nrow(defects); old_count2 <- ndonor(defects)
  defects <- defects %>%
    filter(Rajoite_alkupvm <= threshold)
  cat(sprintf("Defects: Dropped %i / %i donations (%i / %i donors) because Rajoite_alkupvm greater than %s\n",
             old_count - nrow(defects), old_count, old_count2 - ndonor(defects), old_count2, threshold))
  return(defects)
}

```

```{r Load data sets}
filename <- "~/FRCBS/blood_health_phewas/results/phenotype_data.Rdata"
if (! file.exists(filename)) { 
  deferral <- get_deferral()
  donation <- get_donation("~/proj/interval_prediction/data/FRC.DW_DONATION.dat")
  donor <- get_donor("~/proj/interval_prediction/data/FRC.DW_DONOR.dat")
  age <- get_age("~/proj/interval_prediction/data/FRC.DW_DONOR.dat")   # age at date 'threshold'
  finngen <- get_finngen()
  finngen2 <- finngen %>% mutate(smoking = fct_explicit_na(smoking, na_level = NA))
  fd_filename <- "~/proj/interval_prediction/data/DataFromMuriel/20181129/r02ds.donorData.rdata"
  findonor <- load_single(fd_filename)
  findonor <- findonor %>% anti_join(finngen2, by="donor") # There are four donors that are in both datasets.
                                                           # In that case, use the information in FinnGen
  findonor2 <- findonor %>% 
    select(donor, weight, height, smoking=QR54, gender=Gender) %>%
    mutate(smoking = fct_recode(smoking, 
                                `yes, occassionally` = "sometimes",
                                `yes, regularly` = "daily"),
           smoking = factor(smoking, ordered = FALSE),
           smoking = fct_explicit_na(smoking, na_level = NA))
  finngen <- bind_rows(finngen2, findonor2)
  save(list = c("deferral", "donation", "donor", "age", "finngen"), file = filename)
} else {
  load(filename, verbose = TRUE)
}

```

```{r}
create_venn <- function(setlist) {
  plot <- RVenn::ggvenn(RVenn::Venn(setlist), thickness = 1.0, 
                 fill = c("gold", "deeppink", "dodgerblue3"), 
                 slice=names(setlist))
  return(plot)
}
```

Intersections of deferral, finngen, and donation dataframes.

```{r}


# Get intersection of donors
donor_deferral <- deferral %>% pull(donor)
donor_donation <- donation %>% pull(donor)
donor_finngen <- finngen %>% pull(donor)

setlist <- list(deferral=donor_deferral, donation=donor_donation, finngen=donor_finngen)
plot <- create_venn(setlist)
plot

#intersection_donors <- intersect(intersect(donor_deferral, donor_donation), donor_finngen)
intersection_donors <- intersect(donor_donation, donor_finngen)

cat(sprintf("Intersection donors: %i, deferral: %i, donation: %i, FinnGen: %i\n", length(intersection_donors), 
            length(donor_deferral), length(donor_donation), length(donor_finngen)))

# Restrict to donors that are available in both eProgesa and in FinnGen
deferral <- deferral %>% filter(donor %in% intersection_donors)
donation <- donation %>% filter(donor %in% intersection_donors)
finngen  <- finngen %>% filter(donor %in% intersection_donors)
```

## Compute the covariates

```{r Compute covariates}
covariates <- finngen %>% 
  mutate(sex=ifelse(gender=="Women", 1, 0),
         smoking=ifelse(smoking!="no", 1, 0)) %>%
  select(donor, sex, weight, height, smoking)   # Should we compute the age at the end of 2019?
  
covariates <- covariates %>%   # Add the recomputed age
  inner_join(age, by="donor")
```


## Compute the phenotypes

```{r Compute phenotypes}

donor2 <- donor %>% select(donor=KEY_DONOR, sex=KEY_DONOR_SEX)
if (! "sex" %in% names(donation)) {
  donation <- donation %>% inner_join(donor2, by="donor")
}

tmp <- deferral %>% 
  group_by(donor, KEY_DEFERRAL_DATE_START) %>% 
  summarise(S821_or_P821 = length(intersect(KEY_DEFERRAL_CODE, c("S821", "P821"))) >= 1, .groups="drop") %>%
  filter(S821_or_P821 == TRUE) %>% 
  mutate(KEY_DEFERRAL_CODE="S821_or_P821") %>% 
  select(-S821_or_P821) %>%
  left_join(donation, by=c("donor"="donor", "KEY_DEFERRAL_DATE_START"="date"))

# A donor can also be deferred if his/her Hb drops by 20 unit since last visit
S821_or_P821 <- tmp %>% 
  filter(((sex=="Man" & Hb < 135) | (sex=="Woman" & Hb < 125)) | (previous_Hb - Hb > 20)) %>%
  select(donor, KEY_DEFERRAL_CODE, KEY_DEFERRAL_DATE_START)

bad  <- tmp %>% filter(((sex=="Man" & Hb >= 135) | (sex=="Woman" & Hb >= 125) & (previous_Hb - Hb < 20)))
cat(sprintf("In %s deferral cases Hb was above threshold but Hb hadn't drop by at least 20 units", nrow(bad)))                                

hb_stuff <- donation %>% 
  group_by(donor) %>% 
  summarise(hb_median=median(Hb, na.rm=T), hb_mad=mad(Hb, na.rm=T), tries=n()) %>%
  drop_na()


# How should we define a donation here? Should we get it directly from some progesa field?
successful_donations <- donation %>% 
  filter(donat_phleb == "K") 

lifetime_donations <- successful_donations %>% 
  group_by(donor) %>% 
  summarise(lifetime_donations = n())

last_two_years_donations <- successful_donations %>%
  filter(date >= threshold - years(2)) %>%   
  group_by(donor) %>% 
  summarise(last_two_years_donations = n())



# Different deferral categories' fractions
count_S821 <- deferral %>% filter(KEY_DEFERRAL_CODE=="S821") %>% count(donor, name="count_S821")
count_P821 <- deferral %>% filter(KEY_DEFERRAL_CODE=="P821") %>% count(donor, name="count_P821")
count_P820 <- deferral %>% filter(KEY_DEFERRAL_CODE=="P820") %>% count(donor, name="count_P820")
bin_S821 <- count_S821 %>% mutate(bin_S821 = as.integer(count_S821 > 0)) %>% select(donor, bin_S821)
bin_P821 <- count_P821 %>% mutate(bin_P821 = as.integer(count_P821 > 0)) %>% select(donor, bin_P821)
bin_P820 <- count_P820 %>% mutate(bin_P820 = as.integer(count_P820 > 0)) %>% select(donor, bin_P820)

count_S821_or_P821 <- S821_or_P821 %>% 
  filter(KEY_DEFERRAL_CODE=="S821_or_P821") %>% 
  count(donor, name="count_S821_or_P821")
bin_S821_or_P821 <- count_S821_or_P821 %>% 
  mutate(bin_S821_or_P821 = count_S821_or_P821 > 0) %>% 
  select(donor, bin_S821_or_P821)

defects <- get_defects()

# Serious local effects
local_effect_codes <- str_split("H1M0 H1V1 H1VL H1ML H2V0 H2VL H3V0 H3M0 H5V0 H5M0 H3VL H5ML H3ML H4M0 H4ML", " ")[[1]]   # old
#codes1 <- str_split("H01E H01L H02E H02L H03E H03L H04E H04L H05E H05L", " ")[[1]]   # New CRM
#serious_local_ever <- defects %>% filter(Rajoite_koodi %in% codes1) %>% count(donor, name="serious_local_ever")
serious_local_ever <- defects %>% group_by(donor) %>% summarise(serious_local_ever=max(Rajoite_koodi %in% local_effect_codes))

# Fainting
fainting_codes <- str_split("H6V1 H6V3 H6M1 H6M3 H6VL H6ML H6V2", " ")[[1]]   # old
better_fainting_codes <- str_split("H6V1 H6V3 H6VL H6V2", " ")[[1]]   # old # No later faintings here
#codes2 <- str_split("H06L H07E H07L H08E H08L", " ")[[1]]   # New CRM
#fainting_ever <- defects %>% filter(Rajoite_koodi %in% codes2) %>% count(donor, name="fainting_ever")
fainting_ever <- defects %>% group_by(donor) %>% summarise(fainting_ever=max(Rajoite_koodi %in% fainting_codes))

later_fainting <- setdiff(fainting_codes, better_fainting_codes)

compute_better_fainting <- function(donation, defects) {  
  df <- left_join(donation, defects, by=c("donor", "date"="Rajoite_alkupvm")) %>%
    group_by(donor) %>% 
    summarise(better_fainting_ever=max(case_when(
      Rajoite_koodi %in% better_fainting_codes & volume_drawn > 0 ~ 1,
      Rajoite_koodi %in% later_fainting | (Rajoite_koodi %in% better_fainting_codes & volume_drawn <= 0) ~ NA_real_,
      TRUE ~ 0
      )))
  return(df)
}
better_fainting_ever <- compute_better_fainting(donation, defects)

# Get sTfR
filename <- "~/FRCBS/blood_health_phewas/cambridge_collaboration/data/pheno.rds"
stfr <- readRDS(filename)
stfr <- stfr %>% select(donor, stfr)

# Join all results into one dataframe
phenotypes <- hb_stuff
phenotypes <- left_join(phenotypes, lifetime_donations)
phenotypes <- left_join(phenotypes, last_two_years_donations)
phenotypes <- left_join(phenotypes, count_S821)
phenotypes <- left_join(phenotypes, count_P821)
phenotypes <- left_join(phenotypes, count_P820)
phenotypes <- left_join(phenotypes, count_S821_or_P821)
phenotypes <- left_join(phenotypes, bin_S821)
phenotypes <- left_join(phenotypes, bin_P821)
phenotypes <- left_join(phenotypes, bin_P820)
phenotypes <- left_join(phenotypes, bin_S821_or_P821)
phenotypes <- left_join(phenotypes, serious_local_ever)
phenotypes <- left_join(phenotypes, fainting_ever)
phenotypes <- left_join(phenotypes, better_fainting_ever)
phenotypes <- left_join(phenotypes, stfr)


phenotypes <- phenotypes %>% 
  replace_na(list(lifetime_donations=0, last_two_years_donations=0, 
                  count_S821=0, count_P821=0, count_P820=0, count_S821_or_P821=0, 
                  bin_S821=0, bin_P821=0, bin_P820=0, bin_S821_or_P821=0, 
                  serious_local_ever=0, fainting_ever=0))
#, better_fainting_ever=0)) # We allow NA here, they will be dropped later.
```

## Counts of adverse effects in donors.

```{r}
variables <- c("count_S821", "count_P821", "count_P820", "count_S821_or_P821",
               "bin_S821", "bin_P821", "bin_P820", "bin_S821_or_P821",
               "serious_local_ever", "fainting_ever", "better_fainting_ever")
summary(phenotypes %>% transmute_at(variables, as.factor))
```

```{r}
# Convert counts to fractions
phenotypes <- phenotypes %>%
  mutate(S821 = count_S821 / tries,
         P821 = count_P821 / tries,
         P820 = count_P820 / tries
  )
```

## Covariates

```{r Covariate summary}
summary(covariates %>% mutate(across(c(sex, smoking), as.factor)))
```

```{r Covariate plot}
covar_g <- covariates %>% 
  select(-donor) %>% 
  gather() %>% 
  ggplot(aes(value)) + 
  facet_wrap(~key, scale="free") + 
  geom_histogram()
covar_g
```

## Phenotypes

```{r Phenotype summary}
summary(phenotypes %>% mutate(across(c(bin_S821, bin_P821, bin_P820, bin_S821_or_P821, serious_local_ever, fainting_ever, better_fainting_ever), as.factor)))
```

```{r Phenotype descriptions}
phenotype_description <- tibble(
  phenotype=c("hb_median", "hb_mad", "tries", "lifetime_donations", "last_two_years_donations", 
              "S821", "P821", "P820", 
              "count_S821", "count_P821", "count_P820", "count_S821_or_P821",
              "bin_S821", "bin_P821", "bin_P820", "bin_S821_or_P821", 
              "serious_local_ever", "fainting_ever", "better_fainting_ever", "stfr"),
  type=c("continuous", "continuous", "count", "count", "count", 
         "fraction", "fraction", "fraction",
         "count", "count", "count", "count",
         "binary", "binary", "binary", "binary", 
         "binary", "binary", "binary", "continuous"),
  description=c("Lifetime median Hb", "Lifetime mean absolute deviation", "Lifetime donation attempts", "Lifetime donations", "Donation during the last two years", 
                "Fraction of deferrals because Hb < threshold - 10", "Fraction of deferrals because Hb < threshold - 20",
  "Fraction of deferrals because female Hb > 175 or male Hb > 195",
  "Count of S821", "Count of P821", "Count of P820", "Count of S821_or_P821", 
"Any deferrals because Hb < threshold - 10", "Any deferrals because Hb < threshold - 20",
  "Any deferrals because female Hb > 175 or male Hb > 195", "Any S821 or P821 deferrals",
  "Serious local defect ever", "Fainting ever", "Fainting during donation ever",
"Soluble transferrin receptor"))
knitr::kable(phenotype_description)
```


```{r Phenotype plot}
pheno_g <- phenotypes %>% 
  select(-donor) %>% 
  gather() %>% 
  mutate(key=factor(key, levels=phenotype_description$phenotype)) %>%
  ggplot(aes(value)) + 
  facet_wrap(~key, scale="free", ncol=4) + 
  geom_histogram()
pheno_g
```

Correlations of phenotypes.

```{r Correlations}
corrplot::corrplot(cor(phenotypes %>% select(-donor) %>% drop_na(-stfr)))
```

Create sets of donors based on binary variables, and see how the sets overlap.

```{r UpSetR}
create_set <- function(df, variable) {
  df %>% filter(.data[[variable]]==1.0) %>% pull(donor)
}
binary_variables <- c("bin_S821", "bin_P821", "bin_P820", "bin_S821_or_P821", 
                      "serious_local_ever", "fainting_ever", "better_fainting_ever")
#helper(phenotypes, "bin_S821")
sets <- map(binary_variables, function(variable) create_set(phenotypes, variable))
names(sets) <- binary_variables
upset(fromList(sets), order.by = "freq")
```


```{r Write results}
if (write_results) {
  #d <- "2021-08-31"
  d <- Sys.Date()
  ggsave(sprintf("%s/phenotypes-%s.png", result_dir, d), pheno_g)
  ggsave(sprintf("%s/covariates-%s.png", result_dir, d), covar_g)
  write_delim(covariates, sprintf("%s/covariates-%s.txt", result_dir, d), "\t", col_names = TRUE)
  write_delim(phenotypes, sprintf("%s/phenotypes-%s.txt", result_dir, d), "\t", col_names = TRUE)
}

```

## Analyze the fainting cases

How many faintings are there?

```{r}
fainting_defects <- defects %>% 
  filter(Rajoite_koodi %in% fainting_codes) 
kable(fainting_defects %>% 
  count(Rajoite_koodi, Rajoite_selite))
```

Some of the above codes don't seem to be faintings at all! Let's try to find all fainting codes:

```{r}
kable(defects %>% filter(str_detect(Rajoite_selite, "[Pp]yört")) %>% count(Rajoite_koodi, Rajoite_selite))
```

```{r}

better_fainting_defects <- defects %>% 
  filter(Rajoite_koodi %in% better_fainting_codes)
```

How long in days is the "restriction"?

```{r}
unique(better_fainting_defects$Rajoite_loppupvm - better_fainting_defects$Rajoite_alkupvm)
```
Get the explanations of Status and Donat_phleb variables in eProgesa

```{r, message=FALSE}
helper <- function(df) {
  df <- df %>% 
    transmute(a=str_split(...1, fixed("|"))) %>%
    unnest_wider(col=a, names_repair="minimal") %>%
    set_colnames(c("Code", "Explanation")) %>%
    mutate(Explanation=str_remove_all(Explanation, "\""))
  df
}
eProgesa_description_filename <- "~/proj/interval_prediction/data/Tilasto - eProgesa_datakentät_taulukkona (ID 7609).xlsx"
dfs2 <- map(c("Status"="Status", "Type"="Type"), function(sheet) read_excel(eProgesa_description_filename, sheet=sheet, col_names=FALSE))
dfs <- map(dfs2, helper)
kable(dfs$Status)
kable(dfs$Type)
```

How do the donations related to faintings look like? Note that the counts here are much smaller than above
because I have restricted the donations to the FinnGen donors and I have taken only donations before 2019-12-31.

```{r}
better_fainting_donations <- inner_join(donation, better_fainting_defects, by=c("donor", "date"="Rajoite_alkupvm"))

summary(better_fainting_donations %>% select(date, status, donat_phleb, directed, volume_drawn, Hb, Ikä_luovutushetkellä, Ikä_tänään))
```

Counts of defects among FinnGen blood donors before 2019-12-31.

```{r Counts of defects in FinnGen}
kable(better_fainting_donations %>% 
  count(Rajoite_koodi, Rajoite_selite))
```

Show separate summaries for each Rajoite_selite:

```{r}
group_walk(better_fainting_donations %>% group_by(Rajoite_selite), 
           function(x,y) {
             cat(sprintf("\nDefect code: %s\n\n", y$Rajoite_selite))
             print(summary(x %>% select(status,donat_phleb, directed, volume_drawn)))
             } )
```

The volume_drawn in more detail:

```{r}
table(better_fainting_donations %>% select(volume_drawn, Rajoite_selite))
```

## Check what the deferral codes actually mean

```{r}
donor2 <- donor %>% select(donor=KEY_DONOR, sex=KEY_DONOR_SEX)
combined <- donation %>% 
  inner_join(deferral, by=c("donor", "date"="KEY_DEFERRAL_DATE_START")) %>%
  #inner_join(donor2, by="donor") %>%
  filter(KEY_DEFERRAL_CODE %in% c("S821", "P821", "P820"))
```

The Hb limits for the deferral classes seem to be vague!

```{r}
hb_range <-combined %>% 
  drop_na(Hb) %>%  # There are NA values for classes S821 and P821!!!!
  group_by(sex, KEY_DEFERRAL_CODE) %>%
  summarise(min=min(Hb), max=max(Hb), .groups = "drop")
hb_range
```

```{r}
generate_my_breaks <- function(step) {
  # Break between limits at every position that is multiple of 'step' 
  my_breaks <- function(limits) {
    #step <- 0.2
    m <- limits %/% step
    m <- ifelse(m < 0, m+1, m)
    m <- m*step
    return(seq(m[1], m[2], step))
  }
  return(my_breaks)
}

```


```{r}
combined2 <-combined %>% 
  drop_na(Hb) %>%  # There are NA values for classes S821 and P821!!!!
  mutate(class=str_c(KEY_DEFERRAL_CODE, sex, sep = " "))
```

The main blobs seem to have Hb range of 10.

```{r Sina plot of Hb versus deferral class}
limits <- tribble(
  ~class, ~ymin, ~ymax,
  "S821 Woman", 115, 124,
  "S821 Man", 125, 134,
  "P821 Woman", -Inf, 114,
  "P821 Man", -Inf, 124,
  "P820 Woman", 175, Inf,
  "P820 Man", 195, Inf
)
g <- combined2 %>% ggplot(aes(class, Hb)) + 
  ggforce::geom_sina() + 
  geom_errorbar(mapping=aes(x=class, ymin=ymin, ymax=ymax),
                 data=limits,
                 inherit.aes = FALSE, linetype="solid", color="red") +
  scale_y_continuous(breaks = generate_my_breaks(10)) +
  coord_flip()
g
```

Boxplot is less clear than sina plot.

```{r Boxplot of Hb versus deferral class}
combined2 %>% ggplot(aes(class, Hb)) + 
  geom_boxplot() + 
  scale_y_continuous(breaks = generate_my_breaks(10)) +
  coord_flip()
```

What could the limits be?

```{r Barplot of Hb in deferral class range}
combined2 %>% 
  filter((class=="S821 Woman" & 113 <= Hb & Hb <= 127) |
           (class=="S821 Man" & 123 <= Hb & Hb <= 137) |
           (class=="P821 Woman" & 100 <= Hb & Hb <= 117) |
           (class=="P821 Man" & 110 <= Hb & Hb <= 127) |
           (class=="P820 Woman" & 173 <= Hb) |
           (class=="P820 Man" & 193 <= Hb)
  ) %>% 
  mutate(class = fct_relevel(class, c("P821 Man", "S821 Man", "P820 Man", "P821 Woman", "S821 Woman", "P820 Woman"))) %>%
  ggplot(aes(Hb)) + 
  geom_bar() + 
  #facet_grid(KEY_DEFERRAL_CODE ~ sex, scales="free")
  facet_wrap(vars(class), scales="free")
```

So, it would seem that the limits are:
P821: Hb <= 124 (male), Hb <= 114 (female)
S821: 125 <= Hb <= 134 (male), 115 <= Hb <= 124 (female)
P820: inconclusive, but maybe 195 < Hb (male), 175 < Hb (female) 

Donors and days that don't fit the deferral class:

```{r Misclassification as a function of time}
odd_keys <- combined2 %>% 
  filter((class=="S821 Woman" & (Hb <= 114 | 125 <= Hb)) |
           (class=="S821 Man" & (Hb <= 124 | 135 <= Hb)) |
           (class=="P821 Woman" & 115 <= Hb) |
           (class=="P821 Man" & 125 <= Hb)
  ) %>%
  select(donor, date) #%>%
  #distinct()

cat(sprintf("There are %i donations whose Hb does not agree with their deferral class\n", nrow(odd_keys)))

odd_keys %>% ggplot(aes(date)) +
  geom_histogram()
```

```{r}
odd_donor_days <- inner_join(donation, donor2, by="donor") %>% 
  semi_join(odd_keys)

# How many donations per day?
odd_donor_days %>% count(donor, date) %>% pull(n) %>% table
```

Are there multiple donation attempts per day per donor?

```{r}
donation %>% count(donor, date) %>% count(n) %>% rename("Attempts per day"=n, "Donor days"=nn)
```


Are there multiple deferral codes per day and donor in the deferral table?

```{r}
deferral %>% 
  filter(KEY_DEFERRAL_CODE %in% c("S821", "P821", "P820")) %>% 
  group_by(donor, KEY_DEFERRAL_DATE_START) %>% 
  #filter(n() > 1) %>% 
  summarise(codes=as.factor(paste(sort(KEY_DEFERRAL_CODE), collapse=" ")), n=n(), .groups="drop") %>% 
  summary()
```

But even these don't explain all the 813 mislabeled donations!







