---
title: "PCA components of the genetic relationship matrix"
author: "Jarkko Toivonen"
date: "2021-04-07"
output:
  md_document: default
  html_document: default
  pdf_document: default
params:
  input_file: NULL
  recompute_results: FALSE
  result_dir: "/home/toivoja/FRCBS/blood_health_phewas/phewas-30k/results"
#  prefix: "/home/toivoja/FRCBS/blood_health_phewas/results/pca-cleaned/pca-FG-1KG-intersection"
#  prefix: "/home/toivoja/FRCBS/blood_health_phewas/phewas-30k/results/pca-cleaned/pca-FG-1KG-intersection"
  dir_1kg: "~/data/public/1000genomes"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GADMTools)
#library(gridExtra)
#library(ggpubr)
#library(tidyverse)
library(viridis)
library(knitr)
library(RColorBrewer)
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(parallel))
suppressPackageStartupMessages(library(gridExtra))
suppressPackageStartupMessages(library(ggpubr))

# Files from 1000 genomes are here

#prefix=params$prefix
dir_1kg=params$dir_1kg
# This contains the PCA result
result_dir <- params$result_dir
fig_dir <- paste(result_dir, "images", sep="/")
pca_cleaned_dir <- paste(result_dir, "pca-cleaned", sep="/")
eigenvector_filename <- paste(pca_cleaned_dir, "pca-FG-1KG-intersection.eigenvec", sep="/")
eigenvalue_filename  <- paste(pca_cleaned_dir, "pca-FG-1KG-intersection.eigenval", sep="/")
ultimate_eigenvector_filename <- paste(pca_cleaned_dir, "ultimate-pca.eigenvec", sep="/")
ultimate_eigenvalue_filename  <- paste(pca_cleaned_dir, "ultimate-pca.eigenval", sep="/")
outliers_filename             <- paste(fig_dir, "pca_outliers.png", sep="/")
barplot_filename              <- paste(fig_dir, "pca_barplot.png", sep="/")
aberrant_outliers_filename    <- paste(fig_dir, "aberrant_pca_outliers.png", sep="/")
aberrant_results_filename     <- paste(pca_cleaned_dir, "aberrant_results.rdata", sep="/")
save_figs <- FALSE
```




Read PCA data

```{r}
pca <- read_delim(eigenvector_filename, delim=" ", col_names = TRUE)
```




## Plot the principal component pairs of FinnGen blood donors and people from 1000 genomes


Create population labels from the 1000 genomes data and the FinnGen blood donor data.

```{r}
# Get population labels from 1000 Genomes data
population_filename <- paste(dir_1kg, "integrated_call_samples_v3.20130502.ALL.panel", sep="/")
# The input file has stupidly two tab characters at the end of the header
population <- read_delim(population_filename, delim="\t", col_names = c("sample", "pop", "super_pop", "gender"))


# Create population name for each individual.
# Use super population except for Finns use the population (FIN).
population <- population %>% 
  mutate(super_pop=ifelse(pop=="FIN", "FIN", super_pop)) %>%
  select("IID"="sample", "population"="super_pop")

# Create population labels for individuals (samples) in the combined FinnGen and 1KG data sets
ids <- tibble(IID=pca$IID)
population <- left_join(ids, population, by="IID") %>%
  replace_na(list(population="FinnGen")) # Those not in 1kg are in FinnGen
```


```{r create pca_and_population table}


# Join the IID/population dataframe and the PCA dataframes
pca_and_population <- pca %>% inner_join(population, by="IID")


population_names <- unique(c("FIN", "FinnGen", pca_and_population$population))
#temp <- tibble(population=population_names)
#pca <- inner_join(temp, pca, by="population")  # Just to reorder
pca_and_population <- pca_and_population %>% mutate(population = factor(pca_and_population$population, levels=population_names))
#pca <- pca %>% select(everything(), "population")

# This order defines the plotting order
pca_and_population <- arrange(pca_and_population, desc(population))
#print(head(pca))
#summary(pca)
```

```{r plot the principal component pairs}
# plot of the principal components pairs

# Pairs of the first five principal components to be plotted
column.combinations <- combn(1:5, 2)  # Creates a 2x10 matrix of numbers 1,...,5

# Helper to create 'n' evenly spaced color hues as hex codes
gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

# Assign color to each population
n = length(unique(pca_and_population$population))  # Number of populations
colors <- c("#0000FF", "#000000", gg_color_hue(n-2))   # Blue for 1KG FIN and black for FinnGen
names(colors) <- population_names
#print(colors)

# Plot the ith pair of principal components
plot.function <- function(i){
  xi <- column.combinations[1,i]
  yi <- column.combinations[2,i]
  xcolname <- sprintf("PC%i", xi)
  ycolname <- sprintf("PC%i", yi)
  x.label <- paste("Principal Component ", xi)
  y.label <- paste("Principal Component ", yi)

  plot <- pca_and_population %>% 
    ggplot(aes_string(x=xcolname, y=ycolname)) + 
    geom_point(aes(colour = population), size = 2) + #, alpha=0.5) + 
    labs(x = x.label, y = y.label, colour= "Sample") + 
    theme_minimal() + 
    theme(axis.text.y = element_text(angle = 90, size = 6), 
          axis.text.x = element_text(size = 6), 
          axis.title=element_text(size=13)) + 
    scale_color_manual(values=colors)
  
  return(plot)
}

plot_list <- lapply(1:ncol(column.combinations), FUN= plot.function)
pc_plot <- ggpubr::ggarrange(plotlist=plot_list, common.legend=T, ncol=2, nrow=5, legend="bottom", labels = paste(LETTERS[1:length(plot_list)], ")", sep=""))

running_under_knitr <- isTRUE(getOption('knitr.in.progress'))
if (save_figs || running_under_knitr) {
  ggsave(outliers_filename, plot=pc_plot, device="png", height= 40, width= 30, units="cm")
}

if (running_under_knitr) {
  knitr::include_graphics(outliers_filename)
} else {
  plot_list
}
```

Barplot of the explained variances

```{r}

variance <- read_delim(eigenvalue_filename, delim=" ", col_names = FALSE)
print(head(variance))
colnames(variance) <- "X1"
variance <- variance %>%
  mutate(PC = 1:n(),
         plotted = 100 * X1/sum(X1))

ggplot(variance, aes(y=plotted, x=PC)) + 
  geom_bar(stat="identity") + 
  labs(x = "Principal Component", y= "% Variance explained") + 
  scale_x_continuous(breaks = 1:20, labels=1:20) + 
  theme_classic2() + 
  theme(axis.line.x = element_line(colour = "white"), 
        axis.ticks.x = element_blank(), 
        axis.text = element_text(size=12), 
        axis.title=element_text(size=14))
if (save_figs) {
  ggsave(barplot_filename, device="png", height= 20, width= 30, units="cm")
}
```
## Try to find outliers using the Aberrant package

```{r}
#library(aberrant)
```

```{r}
# Plot the ith pair of principal components
aberrant_function <- function(i){
  xi <- column.combinations[1,i]
  yi <- column.combinations[2,i]
  xcolname <- sprintf("PC%i", xi)
  ycolname <- sprintf("PC%i", yi)
  x <- pca_and_population %>% filter(population %in% c("FinnGen")) %>% dplyr::select(all_of(xcolname), all_of(ycolname))
  res <- aberrant::aberrant (x, lambda=30, alpha=1, beta=20, standardize= FALSE, uncorr = TRUE, verbose=FALSE)
  res
}
```

```{r}
if (params$recompute_results || !file.exists(aberrant_results_filename)) {
  future::plan(future::multisession, workers = 4)
  results <- furrr::future_map(1:ncol(column.combinations), aberrant_function, .options = furrr::furrr_options(seed = TRUE))
  save(results, file=aberrant_results_filename)
} else {
  load(aberrant_results_filename, verbose=TRUE)
}
NULL
```

Visualize the results. The ellipse is the 99% confidence area.

Making a combined plot does not seem to work as aberrant.plot does not return anything.

```{r}
aberrant_plots <- map(results, aberrant::aberrant.plot)
combined_aberrant_plot <- ggpubr::ggarrange(plotlist=aberrant_plots, common.legend=T, ncol=2, nrow=5, legend="bottom", labels = paste(LETTERS[1:length(aberrant_plots)], ")", sep=""))

#running_under_knitr <- isTRUE(getOption('knitr.in.progress'))
if (save_figs || running_under_knitr) {
  ggsave(aberrant_outliers_filename, plot=combined_aberrant_plot, device="png", height= 40, width= 30, units="cm")
}

if (FALSE && running_under_knitr) {
  knitr::include_graphics(aberrant_outliers_filename)  # does not work because aberrant.plot does not return anything
} else {
  invisible(aberrant_plots)
}
```

We use the first two principal components to decide the outliers.
Verify that we get the same number of outliers as in the figure.

```{r}
group <- results[[1]]$group
table(group)
```
Get the ids of the outliers.

```{r}
outliers <- pca_and_population %>% filter(population %in% c("FinnGen")) %>% filter(group==1)
outlier_ids <- outliers %>% pull(IID)
if (TRUE) {
  write_tsv(outliers %>% select(FID, IID), file = paste(result_dir, "outliers.tsv", sep="/"), col_names=FALSE)
}
head(outlier_ids)
```
## Median of principal components in each Municipality in Finland

Get the map of Finnish municipalities

```{r}
mymap <- gadm_sf_loadCountries("FIN", level = 4)   # Level 4 contains the municipalities
```

Get information about Finnish post codes.

```{r}
postcode_filename <- "~/proj/interval_prediction/data/posti/PCF_20210406.dat"
col_positions <- c(1, 6, 14, 19, 49, 79, 91, 103, 111, 112, 117, 147, 177, 180, 200, 220)
col_names <- c(
"Tietuetunnus",
"Ajopäivämäärä",
"Postinumero",
"Postinumeron nimi suomeksi",
"Postinumeron nimi ruotsiksi",
"Postinumeron nimen lyhenne suomeksi",
"Postinumeron nimen lyhenne ruotsiksi",
"Voimaantulopäivämäärä",
"Tyyppikoodi",
"Hallinnollisen alueen koodi",
"Hallinnollisen alueen nimi suomeksi",
"Hallinnollisen alueen ruotsiksi",
"Kunnan koodi",
"Kunnan nimi suomeksi",
"Kunnan nimi ruotsiksi",
"Kunnan kielisuhdekoodi"
)
postcodes2 <- read_fwf(postcode_filename, col_positions = fwf_positions(start=col_positions, end=lead(col_positions)-1, col_names=col_names),
                      locale = locale(encoding = "latin1"))
postcodes <- postcodes2 %>% 
  select(postcode="Postinumero", Municipality="Kunnan nimi suomeksi") %>%
  mutate(postcode=as.integer(postcode))
```

Get the municipality of each blood donor from the eProgesa data

```{r}
donors <- read_delim("~/proj/interval_prediction/data/FRC.DW_DONOR.dat", delim="|", col_names = FALSE)
donors <- donors %>% 
  #mutate(Municipalities=str_to_sentence(X10)) %>%
  rename(donor=X1, postcode=X9) %>%
  mutate(postcode=as.integer(postcode)) %>%
  inner_join(postcodes, by="postcode")
bd_data <- donors %>% 
  count(Municipality)
```


### Blood donors


Number of donors in each municipality.

```{r}
kable(head(bd_data))
```

```{r}
summary(bd_data %>% mutate(across(Municipality, as.factor)))
#summary(bd_data)
```
How are the donors distributed among the municipalities?

```{r}
bd_data %>% ggplot(aes(n)) +
  geom_histogram(binwidth = 1) +
  scale_x_log10(labels=function(x) sprintf("%i", x)) +
  labs(x="Number of blood donors", y="Number of municipalities")

```

Read final PCA data

```{r}
ultimate_pca <- read_delim(ultimate_eigenvector_filename, delim=" ", col_names = TRUE)
```



```{r}
pca_and_municipality <- inner_join(donors, ultimate_pca, by=c("donor" = "IID"))# %>% rename(donor=X1, Municipalities=X10)
```

### Genotyped blood donors:

```{r}
summary(pca_and_municipality %>% count(Municipality) %>% mutate(across(Municipality, as.factor)))
#summary(bd_data)
```

How are the genotyped donors distributed among the municipalities?

```{r}
pca_and_municipality %>% count(Municipality) %>% ggplot(aes(n)) +
  geom_histogram(binwidth = 1) +
  scale_x_log10(labels=function(x) sprintf("%i", x)) +
  labs(x="Number of genotyped blood donors", y="Number of municipalities")

```

Plot those municipalities only with at least two donors.

```{r}
# Plot the given PCA component on the Finnish map
color_by_pca <- function(data, component) {
  data3 <- data %>% 
    group_by(Municipality) %>%
    filter(n() >= 2) %>%
    summarise(m=median(get(component)))
  
  mybreaks <- seq(min(data3$m), max(data3$m), length.out=4)
  
  g1 <- data3 %>% ggplot(aes(x=m)) + geom_histogram(bins=30) + labs(x=sprintf("Median %s", component))
  
  #colbreaks = 5
  #col <- viridis(colbreaks+2)[1:(colbreaks+2)]
  number_of_categories=7
  
  #print(head(data3))
  g2 <- choropleth(mymap, 
             as.data.frame(data3),    # Needs to be a data.frame not a tibble!!!
             adm.join = "Municipality",
             value = "m",
             breaks = "quantile",
             steps = number_of_categories,
#             breaks = mybreaks,
#             palette="RdYlGn",
#             palette="GnYlRd",
#             palette=col,
             palette=brewer.pal(number_of_categories, "RdYlGn"),
             #palette="Oranges",
             legend = sprintf("Median %s", component),
             title=sprintf("Median %s in each municipality", component)) + 
    guides(fill = guide_legend(direction="vertical", reverse=FALSE)) + 
    theme(legend.position="right") 
    
  #print(class(g2))
  ggarrange(plotlist=list(g1, g2), common.legend=FALSE, 
            ncol=2, nrow=1, legend="right")  

}
```

Plot the medians of the first five principal components on the Finnish map

```{r}
color_by_pca(pca_and_municipality, "PC1")
color_by_pca(pca_and_municipality, "PC2")
color_by_pca(pca_and_municipality, "PC3")
color_by_pca(pca_and_municipality, "PC4")
color_by_pca(pca_and_municipality, "PC5")
```









